---
title: "Learning ggplot - Part 2 of 3 "
author: "Ritoban Sen"
#date: 2021-08-08
categories: ["R"]
tags: ["R Markdown", "plot", "regression"]
---

### Working with the data
For the purposes of understanding Statistical Transformations in ggplot2 we will be working with the data set '*diamonds*'.

```{r calling the data, warning=F}
library("tidyverse")
library("gridExtra")

ggplot2::diamonds
?diamonds
```
\

#### Description of the data set
\

##### **Prices of over 50,000 round cut diamonds** \
A dataset containing the prices and other attributes of almost 54000 diamonds. The variables are: \
\
1. **price** : price in US dollars (\$326–\$18,823) \
2. **carat** : weight of the diamond (0.2–5.01) \
3. **cut** : quality of the cut (Fair, Good, Very Good, Premium, Ideal) \
4. **color** : diamond colour, from D (best) to J (worst) \
6. **clarity** : a measurement of how clear the diamond is (I1 (worst), SI2, SI1, VS2, VS1, VVS2, VVS1, IF (best)) \
7. **x** :length in mm (0–10.74) \
8. **y** : width in mm (0–58.9) \
9. **z** : depth in mm (0–31.8) \
10. **depth** : total depth percentage = z / mean(x, y) = 2 * z / (x + y) (43–79) \
11. **table** : width of top of diamond relative to widest point (43–95) \

## 6. Statistical Transformations
```{r barchart}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))
```

In the above bar chart, *cut* (x variable) is plotted against *count* (y variable). \
But *count* is not a variable present in the dataset *diamonds*. \
\
the *geom_bar()* has calculated new values to plot. \
\
**Bar charts, Histograms, Frequency Polygons** bin the data and calculates the bin counts - number of data points falling into each bin.\

* **stat** : The algorithm used for calculating new values for a graph is called *stat* , short for *statistical transformation*.

> every *geom* has a default *stat* and every stat has a default *geom*. For example, *geom_bar()* by default uses *stat_count()*. *geom* funtions and *stat* functions can be used interchangeably.

```{r geom_inter_stat}
ggplot(data = diamonds) +
  stat_count(mapping = aes(x = cut))
```

Defining stat explicitly is useful when we want to override the default settings. Using ?stat_count we can see the list of all the variables that *stat_count* calculates.

```{r variables of stat count }
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, y = ..prop.., group = 1))
```

Another example : *stat_summary()*
```{r stat_summary, warning=F}
ggplot(data = diamonds) +
  stat_summary(
    mapping = aes(x = cut , y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```
\

> Q. What is the default geom associated with stat_summary()? How could you rewrite the last plot to use this geom function instead of the stat function? 

The default *geom* associated with *stat_summary()* is *geom_pointrange()*.\
The default *stat* for *geom_pointrange()* is *stat_identity()*. \
\
So the same plot can be made using *geom_pointrange()* but changing the default *stat* from *identity* to *summary*:
```{r}
ggplot(data = diamonds) +
  geom_pointrange(
    mapping = aes(x = cut, y = depth),
    stat = "summary"
  )
```

In the above plot mean is used to calculate the middle points and standard deviation is used to calculate the end points of the lines. \
To make the plot exactly as before we need to specify *fun.min*, *fun.max* and *fun*:

```{r}
ggplot(data = diamonds) +
  geom_pointrange(
    mapping = aes(x = cut, y = depth),
    stat = "summary",
    fun.min = min,
    fun.max = max,
    fun = median
  )
```



> Q. What is the difference between *geom_bar()* and *geom_col()*?

* **geom_bar()** : Takes only x values as argument to the aesthetic function *aes()*. \
Height of the bars are proportional to the number of cases in each group calculated by *stat_count*.

* **geom_col()** : Takes both x and y values as argument to the aesthetic fucntion *aes()*. \
Height of the bars are equal to the corresponding values of the y variables. \
It uses *stat_indentity()*
\

> Q. Most geoms and stats come in pairs that are almost always used in concert. Read through the documentation and make a list of all the pairs. What do they have in common?

List of geoms and stats : <https://jrnold.github.io/r4ds-exercise-solutions/data-visualisation.html#exercise-3.7.3>

The paired geoms and states mostly share the same name and are default in each other's arguments.

> Q. What variables does stat_smooth() compute? What parameters control its behavior?

*stat_smooth()* computes : \
\
i. **y** = predicted value \
ii. **ymin** = lower pointwise confidence interval around the mean \
iii. **ymax** = upper pointwise confidence interval around the mean \
iv. **se** = standard error \

parameters : \
\

## 7. Position Adjustments

* ### *Colouring*

```{r}
ggpl7 = ggplot( data = diamonds ) +
          geom_bar(mapping = aes(x = cut, colour = cut))
ggpl8 = ggplot( data = diamonds ) +
          geom_bar(mapping = aes(x = cut, fill = cut))

grid.arrange(ggpl7,ggpl8,nrow = 2)
```
As you can see, assigning the *aesthetics* : **colour** and **fill** to the variable *cut* introduces colours in the chart the following ways for better representation of the data.
\
\

* ### Assigning a new variable to the *aesthetic* **colour :**
```{r}
ggpl9 = ggplot( data = diamonds ) +
         geom_bar(mapping = aes(x = cut, colour = clarity))
ggpl10 = ggplot( data = diamonds ) +
          geom_bar(mapping = aes(x = cut, fill = clarity))

grid.arrange(ggpl9,ggpl10,nrow=2)
```
Mapping a second variable to the *colour* or *fill* aesthetic stacks each column according to the *variable* (here - *clarity*).
\
Stacking is done by the **position adjustment** specified by the *position* argument. There are 3 other *position* options : \
1. **identity** \
2. **dodge** \
3. **fill** \
```{r positions}
ggplot( data = diamonds) +                  #identity
 geom_bar(  
  mapping = aes(x = cut, fill = clarity),
  position = "identity",
  alpha = 1/5
  )

ggplot( data = diamonds, mapping = aes(x = cut,colour = clarity) ) +      #identity
  geom_bar(
   position = "identity",
   fill = NA,
  ) +
  theme(
        panel.background = element_blank(),
        axis.line = element_line(colour = "black")
  )

ggplot( data = diamonds, mapping = aes(x = cut,fill = clarity)) +         #dodge
 geom_bar( position = "dodge" )

ggplot( data = diamonds, mapping = aes(x = cut, colour = clarity)) +      #dodge
 geom_bar( position = "dodge", fill = NA ) +
  theme(
       panel.background = element_blank(), axis.line = element_line(colour = "black")
  )

ggplot( data = diamonds , mapping = aes(x = cut, fill = clarity)) +        #fill
 geom_bar( position = "fill" )

ggplot( data = diamonds, mapping = aes(x = cut, colour = clarity)) +       #fill
 geom_bar( position = "fill" , fill = NA) +
  theme(
        panel.background = element_blank(),
        axis.line = element_line(colour = "black")  
  )
```

* **Identity**\
*position = "identity"* will place each object exactly where it falls in the context of the graph. Sometimes the segments overlap. In such cases *alpha* needs to be set to a lower value to allow some transparency.\
* **dodge**\
*position = "dodge"* instead of stacking places the objects within the groups directly beside one another. This helps in within group comparisons.\
* **fill**\
*position = "fill"* works like stacking but makes all the bars the same length. This helps in between group comparisons.\
\

