---
title: "Learning ggplot - Part 3 of 3 "
author: "Ritoban Sen"
categories: ["R"]
tags: ["R Markdown", "plot", "regression"]
---

```{r include=F}
library("tidyverse")
library("gridExtra")
```

## 8. Coordinate Systems
The default coordinate system used by ggplot2 is the cartesian coordinate system. However there are instances where modifications to the cartesian coordinate system becomes useful.

* #### Switching the X and Y axis: *coord_flip()*
```{r flip}
ggpl19 = ggplot( data = mpg, mapping = aes(x = class, y = hwy, colour = drv)) +
  geom_boxplot(show.legend = T)

ggpl20 = ggplot( data = mpg, mapping = aes(x = class, y = hwy, colour = drv)) +
  geom_boxplot(show.legend = F) +
  coord_flip()

grid.arrange(ggpl19,ggpl20,ncol=2)
```

**coord_flip()** is used to switch the x and y axes. This is helpful to view horizontal boxplots or when the labellings of X axis are too long.
\
\

* #### Setting proper aspect ratio for maps: *coord_quickmap()*
```{r quickmap}

nz = map_data("nz")

ggplot(data = nz,
                mapping = aes(long,lat,group = group)) +
  geom_polygon(fill = "white", colour = "black")

ggplot(data = nz,
                mapping = aes(long,lat,group = group)) +
  geom_polygon(fill = "white", colour = "black") +
  coord_quickmap()

```

**coord_quickmap()** sets the aspect ratio of spatial data properly.
\
\

* #### Polar Coordinates: 
```{r polar}
ggplot(diamonds, aes(cut,fill = cut)) +
  geom_bar(width = 1) +
  coord_polar() +
  labs(title = "(Coxcomb Chart)")
```

**coord_polar()** changes the cartesian coordinates to polar coordinates.
\
**Coxcomb Charts** can be formed by drawing a bar chart and changing the coordinates to polar coordinates.

> Q. Turn a stacked bar chart into a pie chart using coord_polar.

```{r stacktopie}
ggpl23 = ggplot(diamonds,aes(factor(1), fill = clarity)) +
  geom_bar(width = 1,show.legend = F) +
  coord_polar(theta = "y") +
  labs(title = "Pie Chart")

ggpl24 = ggplot(diamonds, aes(factor(1),fill = clarity)) +
  geom_bar(width = 1) +
  coord_polar() +
  labs(title = "Bull's Eye Chart")

grid.arrange(ggpl23,ggpl24,ncol=2)
```

> Q. What’s the difference between coord_quickmap() and
coord_map()?

> Useful Information:

**coord_fixed()**\
```{r fixed}
ggplot( mpg , aes(cty,hwy)) +
  geom_point() +
  geom_abline() +
  coord_fixed()

ggplot( mpg, aes(cty,hwy)) +
  geom_point() +
  geom_abline()

```

Here, *coord_fixed()* causes the *geom_abline()* to draw a line at a 45 degree angle. Useful for visual analysis.

